Inference using context-dependent models of mutation

### 2017-06-03 writeups/context-dependence.tex
mcmc tree models

In the context of genomics,
it is well-known that certain short nucleotide sequences
are in many organisms much more, or less, abundant than expected by chance \citep{burge1992underrepresentation},
due to the combined effects of context dependence of the nucleotide mutation process,
selective constraints on the function of the sequence,
and other processes such as biased gene conversion \citep{duret2009biased}.
Indeed, molecular studies have demonstrated that the spectrum of new mutations in humans
is highly context dependent \citep{schaibley2013influence}
and that multinucleotide substitutions are relatively common \citep{schrider2011pervasive,terekhanova2013prevalence,harris2013errorprone}.
(We are concerned with effects that are \emph{homogeneous} across the sequence,
so do not consider further the many constraints on protein-coding sequence \citep[reviewed in][]{thorne2007protein}.)

@@

%%%%%%%%%%%
\subsection{Homonid data}

@@

      \item[(repair)] In addition to the above, a collection of motifs associated with known biochemical causes
          curated from the literature:
          deamination due to repair of UV-induced pyramidine dimers ((\nT/\nC)\nC $\to$ (\nT/\nC)\nT and \nC\nC$\to$\nT\nT, \citet{uv_damage});
          an interaction between CpG and pyramidine dimers (\nT\nC\nG$\to$\nT\nT\nG and \nC\nC\nG$\to$\nC\nT\nG, \citet{cpg_review});
          repair of guanine adducts at CpG sites (\nC\nG$\to$\nC\nT, \citet{tobacco});
          deamination by AID (\nW\nR\nC\nY$\to$\nW\nR\nT\nY, \nW is \nA or \nT; \nR is \nA or \nG; \nY is \nT or \nC, \citet{AID});
          deamination by members of the APOBEC family (\nT\nC\nW$\to$\nT\nT\nW or \nT\nC\nW$\to$\nT\nG\nW, \citet{APOBEC});
          and repair by error-prone polymerases pol $\eta$ (\nW\nA$\to$\nW\nG, \citet{pol_eta});
          and pol $\iota$ (\nA\nA$\to$\nG\nT and \nA\nA$\to$\nA\nG \citet{pol_iota}):
          \begin{center}
            \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c|c@{\quad$\to$\quad}c@{\quad at rate\quad }c|c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
                \nT\nC     &   \nT\nT      &  $\mu_{UV1}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
                \nG\nA     &   \nA\nA      &  $\mu_{UV1}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
                \nC\nC     &   \nC\nT      &  $\mu_{UV2}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
                \nG\nG     &   \nA\nG      &  $\mu_{UV2}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
                \nC\nC     &   \nT\nT      &  $\mu_{UV3}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
                \nG\nG     &   \nA\nA      &  $\mu_{UV3}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
                \nT\nC\nG  &   \nT\nT\nG   &  $\mu_{mUV}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
                \nC\nC\nG  &   \nC\nT\nG   &  $\mu_{mUV}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
                \nC\nG\nA  &   \nC\nA\nA   &  $\mu_{mUV}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
                \nC\nG\nG  &   \nC\nA\nG   &  $\mu_{mUV}$ & \nA  &   \nG   &  $\mu_{r}$ & \nA  &   \nG   &  $\mu_{v}$ \\
            \end{tabular}
          \end{center}

@@

>>>>

In the context of genomics,
it is well-known that certain short nucleotide sequences
are in many organisms much more, or less, abundant than expected by chance \citep{burge1992underrepresentation},
due to the combined effects of context dependence of the nucleotide mutation process,
selective constraints on the function of the sequence,
and other processes such as biased gene conversion \citep{duret2009biased}.
Indeed, molecular studies have demonstrated that the spectrum of new mutations in humans
is highly context dependent \citep{schaibley2013influence}
and that multinucleotide substitutions are relatively common \citep{schrider2011pervasive,terekhanova2013prevalence,harris2013errorprone}.
In humans, the molecular processes underlying mutations are best studied in contexts of intense somatic mutation:
cancer and the hypervariable regions that generate immune system diversity \citep{evolution_within},
which provide quite different cellular contexts but share commonalities due to shared molecular machinery.
In general, substitution rates depend on sequence context due to the interaction of a great many factors including:
(a) likelihood of DNA damage or errors in cell division (e.g., UV damage \citep{UV} or non-homologous end joining \citep{NHEJ});
(b) activity levels and structural properties of the enzymes responsible for repair (e.g., base excision repar \citep{BER} and Y-family polymerases \citep{polymerase_review});
and (c) any functional effects of an error (e.g., lethal mutations will not be seen in a population census).
(We are concerned with effects that are \emph{homogeneous} across the sequence,
so do not consider further the many constraints on protein-coding sequence \citep[reviewed in][]{thorne2007protein}.)
Some of these effects -- particularly, activity levels of different enzymes in the germline --
will likely vary over evolutionary time,
and so comparisons between widely separated species will identify time-averaged mutation rates.
Recent studies have used nonnegative matrix factorization to identify sets of mutational patterns
co-occurring in comparisons (a) between different cancer types \citep{cancer_scan} and (b) between different human populations \citep{harris,matthieson},
to disentangle distinct signals presumably coming from the action of distinct sources of mutation.

@@

%%%%%%%%%%%
\subsection{Hominid data}

@@

      \item[(repair)] In addition to the above, a collection of motifs associated with known biochemical causes
          curated from the literature:
          deamination due to repair of UV-induced pyramidine dimers ($\mu_{UV}$: (\nT/\nC)\nC $\to$ (\nT/\nC)\nT and \nC\nC$\to$\nT\nT, \citet{uv_damage});
          an interaction between CpG and pyramidine dimers ($\mu_{mUV}$: \nT\nC\nG$\to$\nT\nT\nG and \nC\nC\nG$\to$\nC\nT\nG, \citet{cpg_review});
          repair of guanine adducts at CpG sites ($\mu_g$: \nC\nG$\to$\nC\nT, \citet{tobacco});
          deamination by AID ($\mu_{AID}$: \nW\nR\nC\nY$\to$\nW\nR\nT\nY, \nW is \nA or \nT; \nR is \nA or \nG; \nY is \nT or \nC, \citet{AID});
          deamination by members of the APOBEC family ($\mu_{APOBEC}$: \nT\nC\nW$\to$\nT\nT\nW or \nT\nC\nW$\to$\nT\nG\nW, \citet{APOBEC});
          and repair by error-prone polymerases pol $\eta$ ($\mu_\eta$: \nT\nA$\to$\nT\nG, \citet{pol_eta});
          and pol $\iota$ ($\mu_{\iota}$: \nA\nA$\to$\nG\nT and \nA\nA$\to$\nA\nG \citet{pol_iota}):
          \begin{center}
            \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }c|c@{\quad$\to$\quad}c@{\quad at rate\quad }c|c@{\quad$\to$\quad}c@{\quad at rate\quad }c}
                \nT\nC     &   \nT\nT      &  $\mu_{UV1}$    & \nC\nG     &   \nC\nT      &  $\mu_{g}$      & \nA\nA\nC\nT  &   \nA\nA\nT\nT   &  $\mu_{AID}$ \\
                \nG\nA     &   \nA\nA      &  $\mu_{UV1}$    & \nC\nG     &   \nA\nG      &  $\mu_{g}$      & \nA\nA\nC\nC  &   \nA\nA\nT\nC   &  $\mu_{AID}$ \\
                \nC\nC     &   \nC\nT      &  $\mu_{UV2}$    & \nT\nC\nA  &   \nT\nT\nA   &  $\mu_{APOBEC}$ & \nA\nG\nC\nT  &   \nA\nG\nT\nT   &  $\mu_{AID}$ \\
                \nG\nG     &   \nA\nG      &  $\mu_{UV2}$    & \nT\nC\nT  &   \nT\nT\nT   &  $\mu_{APOBEC}$ & \nA\nG\nC\nC  &   \nA\nG\nT\nC   &  $\mu_{AID}$ \\
                \nC\nC     &   \nT\nT      &  $\mu_{UV3}$    & \nT\nC\nA  &   \nT\nG\nA   &  $\mu_{APOBEC}$ & \nA\nA\nC\nT  &   \nA\nA\nT\nT   &  $\mu_{AID}$ \\
                \nG\nG     &   \nA\nA      &  $\mu_{UV3}$    & \nT\nC\nT  &   \nT\nG\nT   &  $\mu_{APOBEC}$ & \nA\nA\nC\nC  &   \nA\nA\nT\nC   &  $\mu_{AID}$ \\
                \nT\nC\nG  &   \nT\nT\nG   &  $\mu_{mUV}$    & \nT\nG\nA  &   \nT\nA\nA   &  $\mu_{APOBEC}$ & \nA\nG\nC\nT  &   \nA\nG\nT\nT   &  $\mu_{AID}$ \\
                \nC\nC\nG  &   \nC\nT\nG   &  $\mu_{mUV}$    & \nA\nG\nA  &   \nA\nA\nA   &  $\mu_{APOBEC}$ & \nA\nG\nC\nC  &   \nA\nG\nT\nC   &  $\mu_{AID}$ \\
                \nC\nG\nA  &   \nC\nA\nA   &  $\mu_{mUV}$    & \nT\nG\nA  &   \nT\nC\nA   &  $\mu_{APOBEC}$ & \nT\nA\nC\nT  &   \nT\nA\nT\nT   &  $\mu_{AID}$ \\
                \nC\nG\nG  &   \nC\nA\nG   &  $\mu_{mUV}$    & \nA\nC\nA  &   \nA\nC\nA   &  $\mu_{APOBEC}$ & \nT\nA\nC\nC  &   \nT\nA\nT\nC   &  $\mu_{AID}$ \\
                \nT\nA     &   \nT\nG      &  $\mu_{\eta}$   & \nA\nA    &    \nA\nG      &  $\mu_{\iota1}$ & \nT\nG\nC\nT  &   \nT\nG\nT\nT   &  $\mu_{AID}$ \\
                \nT\nA     &   \nC\nA      &  $\mu_{\eta}$   & \nT\nT    &    \nC\nT      &  $\mu_{\iota1}$ & \nT\nG\nC\nC  &   \nT\nG\nT\nC   &  $\mu_{AID}$ \\
                \nA\nA     &   \nG\nT      &  $\mu_{\iota2}$ &           &                &                 & \nT\nA\nC\nT  &   \nT\nA\nT\nT   &  $\mu_{AID}$ \\
                \nT\nT     &   \nA\nC      &  $\mu_{\iota2}$ &           &                &                 & \nT\nA\nC\nC  &   \nT\nA\nT\nC   &  $\mu_{AID}$ \\
                           &               &                 &           &                &                 & \nT\nG\nC\nT  &   \nT\nG\nT\nT   &  $\mu_{AID}$ \\
                           &               &                 &           &                &                 & \nT\nG\nC\nC  &   \nT\nG\nT\nC   &  $\mu_{AID}$ \\
            \end{tabular}
          \end{center}

@@

In the genomic context,
if we knew the specific mutational spectrum of each possible repair or copying pathways,
then this framework could be used to rigorously estimate their average usages and impacts.
Despite substantial progress over the past 15 years \citep{polymerase_reviews},
this is still far from known.
A promising alternative would be to use the distinct signatures obtained by dimension reduction techniques as in
\citet{cancer_scan}, \citep{matthieson} or \citep{pritchard} as a proxy for unknown but distinct pathways.
